---
title: "nlmixr2utils events"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nlmixr2utils events}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_bus <- exists("rxEventListen", asNamespace("rxode2"))
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = has_bus)
```

```{r no-bus, echo = FALSE, eval = !has_bus, results = "asis"}
cat("> **Note:** the installed rxode2 has no event bus, so the code in this",
    "vignette is shown but not run.\n")
```

This vignette is for authors of driver packages built on nlmixr2utils, such as
nlmixr2scm and nlmixr2sir.  It explains how the worker helper `.plap()`
interacts with the rxode2 event bus, and what a driver has to do so that a
logger records one entry per driver run instead of one per internal fit.

## The event bus in brief

rxode2 has a small publish/subscribe event bus (see `?rxode2::rxEventListen`).
Packages that finish a user-facing operation (a fit, a simulation, a driver
result) call `rxode2::rxEventEmit()`, and listeners registered with
`rxode2::rxEventListen(id, fun)` are called as `fun(event, ...)` with the event
name and the payload fields as named arguments.

The key rule is that events are delivered only at scope depth 0.  Every
emitting function enters a shared "operation scope" (`rxode2::rxEventScope()`,
or `.rxEventEnter()`/`.rxEventExit()`), so only the outermost operation
reports; the `rxSolve()` calls inside a fit and the fits inside a driver are
silent.  The depth is mirrored into the `RXODE2_EVENT_DEPTH` environment
variable so that worker processes *started* inside a scope begin silent too.

The main consumer is [nlmixr2log](https://github.com/nlmixr2/nlmixr2log), but
any package can listen.

## What nlmixr2utils does

nlmixr2utils **emits no events of its own**.  Its only hook is in `.plap()`:
every item is evaluated inside `rxode2::rxEventScope()`, whichever process
runs it:

* in the main process (a `sequential` plan),
* in the `lapply()` fallback when future.apply is not installed, and
* in the workers of a `future` plan.

The last case is why the scope is applied per item rather than relying on the
caller's scope.  A `multisession` (or `cluster`) plan set *before* the driver
is called has workers that were already running when the driver entered its
scope, so they never saw `RXODE2_EVENT_DEPTH` and would otherwise be at depth
0.  Any fit or solve such a worker makes would then emit (in the worker
process, where a listener might be loaded, for example by nlmixr2targets).
Wrapping each item closes that gap.

The scope is looked up when each item runs, in the process that runs it, and
is skipped when that rxode2 has no event bus, so nothing changes with an older
rxode2.  When `rxThreads` is given, the thread setting is applied inside the
scope.  An error in an item still restores the depth.

| Where the item runs                     | Depth inside the item | Events from fits/solves in the item |
|-----------------------------------------|-----------------------|-------------------------------------|
| main process (`sequential` plan)        | caller depth + 1      | none                                |
| `lapply()` fallback                     | caller depth + 1      | none                                |
| worker of a plan set before the call    | at least 1            | none                                |
| plain `future_lapply()` without `.plap` | 0 in such a worker    | emitted (the gap `.plap()` closes)  |

## Seeing it work

A small recording listener stores the name of every event it receives:

```{r listen}
library(nlmixr2utils)
got <- character(0)
rxode2::rxEventListen("demo", function(event, ...) got <<- c(got, event))
```

A tiny model, solved at the top level, reports one `solveComplete` (emitted by
rxode2's `rxSolve()`):

```{r solve, message = FALSE}
mod <- rxode2::rxode2({
  d/dt(central) <- -kel * central
})
solveOne <- function(kel) {
  rxode2::rxSolve(mod, params = c(kel = kel),
                  events = rxode2::et(amt = 100, time = 0) |> rxode2::et(0:4),
                  inits = c(central = 0))
}
invisible(solveOne(0.1))
got
```

The same solves run through `.plap()` are silent, because each item is in the
event scope:

```{r plap}
got <- character(0)
future::plan(future::sequential)
res <- .plap(c(0.1, 0.2, 0.3), function(kel) {
  c(kel = kel, depth = rxode2::rxEventDepth(), n = nrow(solveOne(kel)))
})
do.call(rbind, res)
got
rxode2::rxEventDepth()
```

Inside each item the depth is 1; outside it is back to 0, and no event was
delivered.

## Writing a driver

A driver should report once, on exit, with the result of the whole run.  The
pattern used by `runSCM()` and `runSIR()` is to enter the scope at the start,
and leave it with `.rxEventExit(event, ...)` from `on.exit()`, which lowers the
depth and then emits (so the event is delivered only when the driver itself
was called at the top level).  Every rxode2 call is guarded with an
`exists()` check so the driver still works with an rxode2 that has no bus.

```{r driver}
.hasBus <- function() {
  exists("rxEventEmit", envir = asNamespace("rxode2"), inherits = FALSE)
}

myDriver <- function(kels) {
  if (.hasBus()) getExportedValue("rxode2", ".rxEventEnter")()
  .call <- sys.call()
  on.exit({
    if (.hasBus()) {
      .res <- returnValue()
      .exit <- getExportedValue("rxode2", ".rxEventExit")
      if (is.null(.res)) {
        .exit()            # an error: leave the scope, report nothing
      } else {
        .exit("fitResult", fit = NULL, result = .res, kind = "demo",
              call = .call, fun = "myDriver")
      }
    }
  }, add = TRUE)
  res <- .plap(kels, function(kel) tail(solveOne(kel)$central, 1))
  data.frame(kel = kels, c4 = unlist(res))
}

got <- character(0)
myDriver(c(0.1, 0.2))
got
```

One `fitResult` is delivered for the whole run; none of the solves inside it
report.  A real driver passes its base fit as `fit` (nlmixr2scm and nlmixr2sir
pass the fit given to `runSCM()`/`runSIR()`) and a small summary as `result`;
see the `fitResult` payload in nlmixr2log's documentation.  nlmixr2scm also
emits a `fitComplete` (with `source = "scm"`) for the selected model when it
differs from the base fit.

Two points matter for the scope to stay balanced:

* enter the scope exactly once and leave it exactly once, from `on.exit()`,
  so an error cannot leave the process at a non-zero depth (every later event
  would then be dropped);
* emit through `.rxEventExit()` (or call `rxEventEmit()` only after leaving
  the scope); an event emitted while the driver's own scope is active is
  silently dropped.

## Parallel workers

The same driver behaves identically with a `multisession` plan set before the
call: the workers were started at depth 0, but `.plap()` scopes each item in
the worker, so only the driver's `fitResult` is delivered.  This is not run
here, to keep the vignette fast:

```{r multisession, eval = FALSE}
future::plan(future::multisession, workers = 2)
got <- character(0)
res <- .plap(1:4, function(i) rxode2::rxEventDepth())
unlist(res)    # all >= 1: each item is in the scope in its worker
unlist(future.apply::future_lapply(1:4, function(i) rxode2::rxEventDepth()))
               # all 0: the same workers without .plap()
future::plan(future::sequential)
```

## Cleaning up

```{r unlisten}
rxode2::rxEventUnlisten("demo")
rxode2::rxEventListeners()
```

For full logging of fits, simulations and driver results, use
[nlmixr2log](https://github.com/nlmixr2/nlmixr2log) rather than a hand-written
listener.
